On probabilistic one-vs-one classification

### 2025-03-25 paper.tex
Cleaned up Bayes classifier

\usepackage{multirow}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{rotating}

\newtheorem{prop}{Proposition}

@@

\ShortHeadings{Sample JMLR Paper}{One and Two}
\firstpageno{1}

@@

\title{Comparison of canonical polychotomous coupling methods with the method of Wu-Lin-Weng}

@@

An exact solution to this problem is provided by the Bayes classifier. To simplify the discussion we will assume that $p_i(x) > 0$ for all $x$ and $P_i = 0$ for all $i$. Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines The Bayes classifier predicts that the probability of $x$ belonging to the $i$-th class is
\begin{align}
 p^\Pi_i = \frac{p_i(x) P_i}{\sum_{k=1}^K p_k(x)P_k}.
\end{align}

@@

\begin{align}
	r_{ij}(x) = \frac{p_i(x) P_1}{p_i(x)P_1 + p_j(x)P_2}.
\end{align}

Knowledge of vector $\boldsymbol{p}^\Pi$ is essentially equivalent to knowing the function $r_{ij}(x)$. Indeed, if $p_{i}> 0$ for all $x$, then we have
\begin{align}
r_{ij}(x) = \frac{p_i^\Pi}{p_i^\Pi + p_j^\Pi}. \label{eq:bt1}
\end{align}

Conversely, it is easy to show that  the system of equations \label{eq:bt1} has a unique solution.

\begin{prop} \label{prop:binary2multi}
The system of linear equations \eqref{eq:bt1}  has a unique solution which is the multi-class Bayes classifier.
\end{prop}
\begin{proof}
Indeed, we have 
\begin{align*}
\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
\end{align*}
\end{proof}

@@

Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
\begin{equation}
r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}. \label{eq:bt1}
\end{equation}

The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.

Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	\hat{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

 One should note that 
\begin{itemize}
\item the resulting system of equations will usually not be consistent, because $\hat{r}_{ij}$ are only estimates of true values $r_{ij}$,
\item when probabilities are parametrized in other ways (e.g. other common parametrizations are as odds, or log-odds), the system of equations \eqref{eq:bt} is non-linear.
\end{itemize}

\subsection{Exact desiderata on coupling method}

Let us outline what kind of conditions may be desirable in a coupling method.

@@

\emph{Bayes covariance} is a notion introduced in work \cite{BarredaSuch}. It refers to the behavior of a probabilistic multi-class method built in one-vs-one fashion, when the priors change. Any Bayes classifier changes its prediction when the priors change. The change happens according to Bayes theorem.

@@

@@

A natural way to find  $\hat{\boldsymbol{p}}$ is thus to minimize a functional incorporating divergence measures between these distributions
\begin{align*}
\hat{\boldsymbol{p}} = \argmax_{\boldsymbol{p}} \sum_{ij} w_{ij} \divm (\hat D_{ij}^\textrm{multi}, \hat D_{ij}^\textrm{binary})
\end{align*}

@@

Kullback-Leibler divergence strongly penalizes  mispredictions when the true class is predicted to have very low probability. In the context of classification, when we are primarily concerned with accuracy , one could advantageously use other Bregman divergence derived from a proper score.

For instance, a popular alternative is divergence arising from the quadratic  score (also called Brier score).  This divergence takes the form
$$
\divm\nolimits_\textrm{quad} (\boldsymbol{p}, \boldsymbol{q})= \sum (p_i - q_i)^2.
$$
Unlike Kullback-Leibler divergence, the penalty for incorrect prediction with arbitrarily low probability is bounded (by one). This quadratic score is the basis of the popular Wu-Lin-Weng's method \cite{wu2004probability}.

@@

\subsection{Arboreal-based methods}

@@

\subsection{Normal coupling}

@@

\begin{table}[!ht]
\begin{tabular}{cm{2.5cm}cccm{1.5cm}m{1.5cm}}
&property & Wu-Lin-Weng & radial & normal & arboreal & Hinton's oracle \\
\hline 
\multirow{4}{*}{\begin{turn}{90}\makecell{exact}\end{turn}}
&uniqueness & yes & yes & yes & yes & yes \\
&Bradley-Terry consistency & yes & yes & yes & yes & yes \\
&canonicity & no & yes & yes & no & yes \\
&symmetry & yes & yes & yes & no & yes \\
\hline
\multirow{5}{*}{\begin{turn}{90}\makecell{non-exact}\end{turn}}
&Hinton's minimality & ?  & ++ & ?  & ? & +++ \\
&lack of parameters & yes & yes & yes & yes & yes \\
& amounts to a linear system & yes & yes & yes & yes & yes\\
& linear system of special form & no & yes & yes & yes & yes \\
& iterative method & yes & yes & no & not needed & not needed\\
\hline
\end{tabular}
\caption{Summary of properties of coupling methods described in Section \ref{sec:coupling}.}
\label{tab:summaryCoupling}
\end{table}

@@

Here is a citation \cite{chow:68}.

@@

\appendix
\section{}
\label{app:theorem}

@@

>>>>

\usepackage{multirow}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{tikz-cd}

\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}

@@

\ShortHeadings{Probabilistic OVO}{Probabilistic OVO}
\firstpageno{1}

@@

\title{On probabilistic one-vs-one classification}

@@

\subsection{Bayes classifier}

An exact solution to this problem is provided by the Bayes classifier. To simplify the discussion we will assume that $p_i(x) > 0$ for all $x$ and $P_i > 0$ for all $i$. Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines.

 The Bayes classifier predicts that the probability of $x$ belonging to the $i$-th class is
\begin{align}
 p^\Pi_i(x) = \frac{p_i(x) P_i}{\sum_{k=1}^K p_k(x)P_k}.
\end{align}

@@

\begin{align}
	p_{ij}^\Pi(x) = \frac{p_i(x) P_1}{p_i(x)P_1 + p_j(x)P_2}.
\end{align}

Knowledge of vector $\boldsymbol{p}^\Pi$ is  equivalent to knowing the functions $p^\Pi_{ij}(x)$. Indeed, if $p_{i}> 0$ for all $x$, then we have
\begin{align}
p_{ij}^\Pi(x) = \frac{p_i^\Pi(x)}{p_i^\Pi(x) + p_j^\Pi(x)}. \label{eq:bt1}
\end{align}

Conversely, it is easy to show that  the system of equations \eqref{eq:bt1} has a unique solution.

\begin{prop} \label{prop:binary2multi}
The system of  equations \eqref{eq:bt1}  has a unique solution which is the multi-class Bayes classifier.
\end{prop}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

@@

In practice one does not have know true probability distributions and hence cannot construct Bayes classifier. However, given a sample from the distribution, we can train binary classifiers to provide approximations $r_{ij}(x)$ to pairwise Bayes classifiers $p_{ij}(x)$. The goal of probabilistic one-vs-one classification is to deduce for any $x$ in the feature space $X$ an estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\Pi(x)$. One does it by finding approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

% One should note that 
%\begin{itemize}
%\item the resulting system of equations will usually not be consistent, because $\hat{r}_{ij}$ are only estimates of true values $r_{ij}$,
%\item when probabilities are parametrized in other ways (e.g. other common parametrizations are as odds, or log-odds), the system of equations \eqref{eq:bt} is non-linear.
%\end{itemize}

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix with off-diagonal entries being $r_{ij}(x)$.

Let us outline what kind of conditions may be desirable in a coupling method. In Section \ref{sec:des:exact} we will describe exact mathematical criteria, while in \ref{sec:des:inexact} we shall describe more flexible properties.

\subsection{Exact desiderata on coupling method} \label{sec:des:exact}

@@

\emph{Bayes covariance} is a notion introduced in  \cite{vsuch2016bayes}. It refers to the behavior of a probabilistic multi-class method built in one-vs-one fashion, when the priors change. Any Bayes classifier, whether binary or multiclass changes its prediction when the priors change. Let us describe the process. Suppose we change priors from $\Pi = (P_1, \ldots, P_k)$ to $\Pi'= (P'_1, \ldots, P'_K)$. Then the prediction of Bayes classifiers changes from $\boldsymbol{p}^\Pi= (p_1, \ldots, p_K)$ to the vector 
\begin{align}
\boldsymbol{p}^{\Pi'} \propto (p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}). \label{eq:changePrior}
\end{align}

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $R$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $R^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $V$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	R \arrow[r, "\pi"] \arrow[d, "V"] & R^\pi \arrow[d, "V"] \\
	V(R) \arrow[r, "\pi"] & V(R^\pi) = V(R)^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $V$}
\label{fig:bc}
\end{figure}

@@

\label{sec:des:inexact}

@@

A natural way to find  $\hat{\boldsymbol{p}}$ is thus to minimize a functional incorporating divergence measures between these distributions
\begin{align*}
\hat{\boldsymbol{p}} = \argmin_{\boldsymbol{p}} \sum_{ij} w_{ij} \divm (\hat D_{ij}^\textrm{multi}, \hat D_{ij}^\textrm{binary})
\end{align*}

@@

Kullback-Leibler divergence strongly penalizes  mispredictions when the true class is predicted to have very low probability. In the context of classification, when we are primarily concerned with accuracy , one could advantageously use other Bregman divergence derived from a proper score \cite{gneiting2007strictly}.

For instance, a popular alternative is divergence arising from the quadratic  score (also known as Brier score) \cite{gneiting2007strictly}. This divergence takes the form
$$
\divm\nolimits_\textrm{quad} (\boldsymbol{p}, \boldsymbol{q})= \sum (p_i - q_i)^2.
$$
Unlike Kullback-Leibler divergence, the penalty for incorrect prediction with arbitrarily low probability is bounded (by one). This quadratic score is the basis of the popular Wu-Lin-Weng's method \cite{wu2004probability}.

@@

\subsection{Arboreal methods}

@@

\subsection{Bayes covariant methods}

There are many Bayes covariant methods. For instance any arboreal method is Bayes covariant. Given two Bayes covariant methods one can interpolate between them using their ``linear combination'' - an ensemble which is also Bayes covariant. Let us make it explicit.

 Suppose real numbers $\alpha, \beta$ are given such $\alpha + \beta = 1$. If the prediction of the two Bayes covariant methods $\boldsymbol{B}$ and $\boldsymbol{\tilde B}$ are
\begin{align*}
	\boldsymbol{B}(x) = (p_1, \ldots, p_K)\quad\textrm{and}\quad \boldsymbol{\tilde B}(x) = (\tilde p_1, \ldots, \tilde p_K)
\end{align*}
then the prediction of the ensemble $\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}$ is given by
\begin{align*}
	\bigl(\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}\bigr)(x)\propto (\,p_1^\alpha \tilde p_1^\beta, \ldots, p_K^\alpha \tilde p_K^\beta).
\end{align*}

As an example, consider the classification problem with three classes. We have three arboreal classifiers $\boldsymbol{T}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$ which correspond to star graphs centered respectively at classes $C_1, C_2, C_3$. The classifier 
\begin{align}
\frac13 \boldsymbol{T}_1 \oplus \frac13 \boldsymbol{T}_2 \oplus \frac 13 \boldsymbol{T}_3 = \frac13 \boldsymbol{T}_1 \oplus \frac23 \biggl(\frac12 \boldsymbol{T}_2 \oplus \frac 12 \boldsymbol{T}_3
\biggr), \label{eq:bc1}
\end{align}
is Bayes covariant \emph{and} symmetric. A key result proved in \cite{vsuch2016bayes} is that it is a unique such classifier for $K=3$.

\begin{thm}
	There exists a unique symmetric Bayes covariant coupling methods for $K=3$.
\end{thm}

Let us point out one consequence of the theorem. We could define another ensemble $\boldsymbol{E}$ by averaging the predictions of $\boldsymbol{T}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$:
\[
\boldsymbol{E} = \frac13 \boldsymbol{T}_1(x) + \frac13 \boldsymbol{T}_2(x) + \frac 13 \boldsymbol{T}_3(x).
\]
Note that here we use the plus symbol to denote the vector addition of vectors representing $K$-class probability distributions. This is a different classifier from the one given by  \eqref{eq:bc1} (see Appendix \ref{app:explicit}), and since it is symmetric, it cannot be Bayes covariant.

@@

\begin{table}[!ht]
\begin{tabular}{cm{2.5cm}m{1.5cm}m{1.5cm}ccm{1.5cm}m{1.5cm}}
&property & Hastie-Tibshirani & Wu-Lin-Weng & radial & normal & arboreal & Hinton's oracle \\
\hline 
\multirow{5}{*}{\begin{turn}{90}\makecell{exact}\end{turn}}
&uniqueness &  yes & yes & yes & yes & yes & yes \\
&Bradley-Terry consistency & yes & yes & yes & yes & yes & yes \\
&canonicity & yes & no & yes & yes & no & yes \\
&symmetry & yes & yes & yes & yes & no & yes \\
& Bayes covariant & no & no & no & yes & yes & yes \\
\hline
\multirow{5}{*}{\begin{turn}{90}\makecell{non-exact}\end{turn}}
&Hinton's minimality & ? & ?  & ++ & ?  & ? & +++ \\
&lack of parameters & yes & yes & yes & yes & yes & yes \\
& amounts to a linear system & no & yes & yes & yes & yes & yes\\
& linear system of special form & no & no & yes & yes & yes & yes \\
& iterative method & yes & yes & yes & no & not needed & not needed\\
\hline
\end{tabular}
\caption{Summary of properties of coupling methods described in Section \ref{sec:coupling}.}
\label{tab:summaryCoupling}
\end{table}

\section{Experimental results.}

@@

%Here is a citation \cite{chow:68}.

@@

\appendix
\section{}
\label{app:explicit}

@@

In this appendix we provide explicit predictions of coupling methods used in the paper.
